---
title: "Readme"
output: github_document
date: "2023-04-26"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesTraitR

<!-- badges: start -->
<!-- badges: end -->

The goal of BayesTraitR is to facilitate phylogenetic comparative analyses in BayesTraits for R users. It comprises a suite of functions and tools that will be useful for wrangling datasets and trees into the appropriate format for analyses within BayesTraits as well as for interpreting and post-processing the outcomes of such analyses.


## Installation

You can install the development version of BayesTraitR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("joannabaker/BayesTraitR")
```

You can download the latest version of [BayesTraits](http://www.evolution.reading.ac.uk/BayesTraitsV4.0.1/BayesTraitsV4.0.1.html) from the University of Reading website. Note that this package has been developed with the output from BayesTraitsV4.0.1 and different versions of the program may produce incompatible outputs.


## Example Data

This package contains sample brain and body size data for marsupials, taken from the forthcoming publication:

Venditti C, Baker J & Barton, R (2023) *Co-evolutionary dynamics of mammalian brain and body size*, Currently **in review**.


```{r, echo=FALSE, message =FALSE, fig.cap = "Figure 1. Marsupial Brain and Body Size. There is a clear (and unsurprising) strong positive relationship between brain and body mass in marsupials."}
marsupials = read.table("data-raw/marsupials.txt", sep = "\t", header = T, stringsAsFactors = F)
plot(marsupials$Br~marsupials$Bd, xlab = expression(log[10]~Body~Mass~(g)), ylab = expression(log[10]~Brain~Mass~(g)), pch = 16, col = "red", las = 1, bty = "l")
```

